---
title: "README"
output: 
  md_document:
    variant: markdown_github
---

[![Travis-CI Build Status](https://travis-ci.org/jepusto/SingleCaseES.svg?branch=master)](https://travis-ci.org/jepusto/SingleCaseES)
[![Coverage Status](https://img.shields.io/codecov/c/github/jepusto/SingleCaseES/master.svg)](https://codecov.io/github/jepusto/SingleCaseES?branch=master)
[![](http://www.r-pkg.org/badges/version/SingleCaseES)](https://CRAN.R-project.org/package=SingleCaseES)

# SingleCaseES: A calculator for single-case effect size indices

This package provides R functions for calculating basic effect size indices for single-case designs, including several non-overlap measures and parametric effect size measures, and for estimating the gradual effects model developed by [Swan and Pustejovsky (2018)](https://doi.org/10.1080/00273171.2018.1466681). Standard errors and confidence intervals are provided for the subset of effect sizes indices with known sampling distributions. However, it is important to note that all of the standard errors and confidence intervals are based on the assumption that the outcome measurements are mutually independent.

The available __non-overlap indices__ are:

* Percentage of non-overlapping data (PND)
* Percentage of all non-overlapping data (PAND)
* Robust improvement rate difference (IRD)
* Percentage exceeding the median (PEM)
* Non-overlap of all pairs (NAP)
* Tau non-overlap (Tau)
* Tau-U (including baseline trend adjustment)

The available __parametric effect sizes__ are:

* Within-case standardized mean difference
* Log response ratio (decreasing and increasing)
* Log odds ratio
* The gradual effects model, which can be used to estimate log response ratios or log odds ratios in the presence of time trends during treatment and return-to-baseline phases.

The package also includes two graphical user interfaces (designed using [Shiny](https://shiny.rstudio.com/)) for interactive use, both of which are also available as web apps hosted through [shinyapps.io](https://www.shinyapps.io/): 

* `SCD_effect_sizes()` opens an interactive calculator for the basic non-overlap indices and parametric effect sizes. It is also available at https://jepusto.shinyapps.io/SCD-effect-sizes
* `shine_gem_scd()` opens an interactive calculator for the gradual effects model. It is also available at https://jepusto.shinyapps.io/gem-scd

__*Please note that the web apps should only be used for demonstration purposes*__. For research purposes, please install the R package and run the GUI through Rstudio.

# Installation

The package is currently only available on Github. To install it, you will first need to [install R](http://cran.r-project.org/) and [RStudio](http://www.rstudio.com/products/rstudio/download/). Windows users will also need to [install Rtools](http://cran.r-project.org/bin/windows/Rtools/). All of these programs are freely availble. 

Once you have these programs installed, run the following commands at the RStudio console prompt:

```{r, eval=FALSE}
install.packages("devtools")
install.packages("sourcetools")
install.packages("shiny")
install.packages("markdown")
install.packages("ggplot2")
devtools::install_github("jepusto/SingleCaseES")
```


# Basic usage

After installing the package, you are ready to begin using it. Every time you open a fresh R/RStudio session, you will need to start by loading the package functionality as follows:

```{r}
library(SingleCaseES)
```

The functions for calculating the non-overlap indices and parametric effect size estimates all have the same basic structure. Each function takes two inputs: a vector of outcome values for the baseline phase and a vector of outcome values for the treatment phase. For example, the following code creates an object `A` containing outcomes from a baseline phase an an object `B` containing outcomes from a treatment phase:

```{r}
A <- c(20, 20, 26, 25, 22, 23)
B <- c(28, 25, 24, 27, 30, 30, 29)
```

## NAP 

After inputing the data, non-overlap of all pairs can be calculated as follows:

```{r}
NAP(A, B, improvement = "increase")
```

Because formulas are available for NAP, the function reports the standard error and confidence interval in addition to the point estimate of NAP. Note that these formulas assume that the outcome measurements are mutually independent. In the presence of positive auto-correlation, they may under-estimate the true sampling variability of NAP. 

The `NAP` function assumes that increases in the value of the outcome represent therapeutic improvements. For outcomes where decrease is therapeutically desirable, set the optional argument `improvement` to `"decrease"`:

```{r}
NAP(A, B, improvement = "decrease")
```

Type `?NAP` at the console prompt for further information about the available options for NAP. 

## Other non-overlap indices 

Other effect size calculation functions work very similarly. For example, the robust improvement rate difference can be calculated as follows:

```{r}
IRD(A, B, improvement = "increase")
```

Note that only a point estimate is returned because (to the author's knowledge) there are not available methods for estimating the standard error or confidence interval of IRD. 

## Parametric effect size indices

The function for calculating the within-case standardized mean difference has similar arguments to the non-overlap indices:
```{r}
SMD(A, B)
```

By default, the estimate is calculated using the standard deviation of the baseline phase data only. To use the standard deviation pooled across both phases, set the `std_dev` option to `"pool"`:

```{r}
SMD(A, B, std_dev = "pool")
```

By default, the estimate is calculated using the small-sample correction proposed by Hedges (1981) and often referred to as "Hedges' _g_." The standard error and approximate confidence interval for the standardized mean difference estimate are based on the assumption that the outcome measurements are mutually independent.

The package can also calculate both versions of the log response ratio. The log response ratio-increasing (LRRi) is defined so that therapeutic improvements correspond to positive values. It can be calculated as follows:

```{r}
LRRi(A, B, scale = "count")
```

The log response ratio-decreasing (LRRd) is defined so that therapeutic improvements correspond to negative values. The outcome data in the example is defined so that increases are therapeutic improvements, so we will need to specify the direction of improvement:

```{r}
LRRd(A, B, scale = "count", improvement = "increase")
```

In this case, the LRRi and LRRd estimates have the same magnitude but opposite sign. This will be true for outcomes measured as frequency counts or rates. For outcomes measured as percentages or proportions, LRRd and LRRi will differ in both sign and magnitude. See Pustejovsky (2018) for further details and recommendations about how to choose between LRRi and LRRd.

By default, the LRRi and LRRd estimates are calculated using a small-sample correction proposed by Pustejovsky (2015). See `?LRR` for further details. The standard error and approximate confidence interval for the log response ratio estimate are based on the assumption that the outcome measurements are mutually independent.

## Multiple effect sizes for a single series

The `calc_ES()` function can be used to calculate multiple effect size indices for a single series with A and B phases: 

```{r}
calc_ES(A, B, ES = c("LRRd","LRRi","SMD","NAP"), improvement = "increase")
```

Any optional arguments supplied are applied to the effect sizes for which they are relevant.  

## Graphical user interface

To use the graphical user interface for basic effect sizes, you must first ensure that the `SingleCaseES` package is installed (following the directions above). To start the calculator, type the following commands at the RStudio console prompt:

```{r, eval = FALSE}
library(SingleCaseES)
SCD_effect_sizes()
```

The calculator should then open in your default web browser. 
 
To exit the calculator, close the window in which it appears and click the red "Stop" icon in the upper right-hand corner of the RStudio console window.

# Gradual effects model

## Graphical user interface

To use the graphical user interface for the gradual effects model, type the following commands at the RStudio console prompt:

```{r, eval = FALSE}
library(SingleCaseES)
shine_gem_scd()
```

The calculator should then open in your default web browser. 
 
To exit the calculator, close the window in which it appears and click the red "Stop" icon in the upper right-hand corner of the RStudio console window.

# Citations

Please cite this R package as follows:

> Pustejovsky, J. E. & Swan, D. M. (2018). SingleCaseES: A calculator for single-case effect size indices. R package version `r packageVersion("SingleCaseES")`. Retrieved from https://github.com/jepusto/SingleCaseES

Please cite the web applications as follows:

> Pustejovsky, J. E. & Swan, D. M. (2018). Single-case effect size calculator (Version 0.4.0) [Web application]. Retrieved from https://jepusto.shinyapps.io/SCD-effect-sizes

> Swan, D. M. & Pustejovsky, J. E. (2017). gem_scd: A web-based calculator for the Gradual Effects Model (Version 0.1.0) [Web application]. Retrieved from: https://jepusto.shinyapps.io/gem-scd
